---
title: "01 Eligible Patient Diagnosis Identification"
author: "Uriah Martin-Velez and Rachel Baccile"
---
## Imports and Constants

```{r setup}
library(readr)
library(dplyr)
library(stringr)
library(lubridate)

capricorn_dir <- "Z:/CAPriCORN_raw_ds/CAPriCORN" # Read Data
output_folder <- '//cri-rss.cri.uchicago.edu/chess-lab/staff/rbaccile/Matt/output/'# Write Outputs
```

# Identify inclusion and exclusion codes
```{r}
icd10_snomed_crosswalk <- read.csv(paste0(table_location, "icd10_snomed_crosswalk.csv"))
                                   
snomed_codes <- icd10_snomed_crosswalk$snomed_concept_id

valid_hypo_codes <- c(
  "244.0", "244.1", "244.2", "244.3", "244.8", "244.9", "245.2",
  "E03.2", "E03.3", "E03.4", "E03.5", "E03.8", "E03.9",
  "E06.3", "E06.5", "E89.0"
)
hypo_snomed_codes <- icd10_snomed_crosswalk %>% filter(domain=="Hypothyroidism - inclusion")
hypo_snomed_codes <- hypo_snomed_codes$snomed_concept_id
valid_hypo_codes <- c(valid_hypo_codes, hypo_snomed_codes)

exclusion_snomed_codes <- icd10_snomed_crosswalk %>% filter(domain=="Exclusion")
exclusion_snomed_codes <- exclusion_snomed_codes$snomed_concept_id

exclusion_icd10_codes <- icd10_snomed_crosswalk %>% filter(domain=="Exclusion")
exclusion_icd10_codes <- exclusion_icd10_codes$icd10_target

exclusion_icd9_codes <- c("193", "242.00", "242.01", "242.1", "242.11", "242.2", "242.21", "242.3", "242.31",
                          "242.4", "242.41","242.8", "242.81", "242.9", "242.91")

exclusion_codes <- c(exclusion_icd10_codes, exclusion_icd9_codes, exclusion_snomed_codes)

```

## Qualifying Hypothyroidism Codes from `Diagnosis.csv`

```{r}
hypo_diagnosis <- read_csv(
  file.path(capricorn_dir, "diagnosis.csv"),
  col_select = c(STUDY_ID, DX, DX_TYPE, DX_DATE),
  show_col_types = FALSE
) %>%
  mutate(
    DX = str_to_upper(str_trim(DX)),
    DX_DATE = as.Date(DX_DATE)
  ) %>%
  filter(
    DX %in% valid_hypo_codes
  ) %>%
  transmute(
    cap_id = as.character(STUDY_ID),
    hypo_dx = DX,
    hypo_dx_date = DX_DATE
  )
head(hypo_diagnosis)
```

## Qualifying Hypothyroidism Codes from `Condition.csv`

```{r}
hypo_condition <- read_csv(
  file.path(capricorn_dir, "Condition.csv"),
  col_select = c(STUDY_ID, CONDITION, CONDITION_TYPE, REPORT_DATE, ONSET_DATE),
  show_col_types = FALSE
) %>%
  mutate(
    CONDITION = str_to_upper(str_trim(CONDITION)),
  ) %>%
  filter(
    CONDITION  %in% valid_hypo_codes
  ) %>%
  transmute(
    cap_id = as.character(STUDY_ID),
    hypo_dx = CONDITION,
    REPORT_DATE = REPORT_DATE,
    ONSET_DATE = ONSET_DATE
  )

head(hypo_condition)
```

## Combine and Clean Sources

```{r}
library(lubridate)

hypo_condition$REPORT_DATE <- as.Date(
  parse_date_time(hypo_condition$REPORT_DATE, orders = c("ymd", "mdy"))
)
hypo_condition$ONSET_DATE <- as.Date(hypo_condition$ONSET_DATE)
hypo_condition$hypo_dx_date <- pmin(hypo_condition$ONSET_DATE, hypo_condition$REPORT_DATE, na.rm = TRUE)
hypo_condition <- hypo_condition %>% select(cap_id, hypo_dx, hypo_dx_date) %>% distinct()

hypothyroidism_patients <- bind_rows(
  hypo_diagnosis,
  hypo_condition
) %>%
  filter(
    !is.na(cap_id),
    !is.na(hypo_dx),
    !is.na(hypo_dx_date),
    hypo_dx!="428165003" # Exclude active pregnancy code
  ) %>%
  arrange(cap_id, hypo_dx_date) %>%
  group_by(cap_id) %>%
  slice(1) %>%
  ungroup()

head(hypothyroidism_patients)
```

## Exclusion Codes from `Diagnosis.csv`

```{r}
exclusion_diagnosis <- read_csv(
  file.path(capricorn_dir, "diagnosis.csv"),
  col_select = c(STUDY_ID, DX, DX_TYPE, DX_DATE),
  show_col_types = FALSE
) %>%
  mutate(
    DX = str_to_upper(str_trim(DX)),
    DX_DATE = as.Date(DX_DATE)
  ) %>%
  filter(
    DX %in% exclusion_codes
  ) %>%
  transmute(
    cap_id = as.character(STUDY_ID),
    exclusion_dx = DX,
    exclusion_dx_date = DX_DATE
  )
head(exclusion_diagnosis)
```

## Exclusion Codes from `Condition.csv`

```{r}
exclusion_condition <- read_csv(
  file.path(capricorn_dir, "Condition.csv"),
  col_select = c(STUDY_ID, CONDITION, CONDITION_TYPE, REPORT_DATE, ONSET_DATE),
  show_col_types = FALSE
) %>%
  mutate(
    CONDITION = str_to_upper(str_trim(CONDITION)),
  ) %>%
  filter(
    CONDITION  %in% exclusion_codes
  ) %>%
  transmute(
    cap_id = as.character(STUDY_ID),
    exclusion_dx = CONDITION,
    REPORT_DATE = REPORT_DATE,
    ONSET_DATE = ONSET_DATE
  )

head(exclusion_condition)
```

## Combine and Clean Sources

```{r}
exclusion_condition$REPORT_DATE <- as.Date(
  parse_date_time(exclusion_condition$REPORT_DATE, orders = c("ymd", "mdy"))
)
exclusion_condition$ONSET_DATE <- as.Date(exclusion_condition$ONSET_DATE)
exclusion_condition$exclusion_dx_date <- pmin(exclusion_condition$ONSET_DATE, exclusion_condition$REPORT_DATE, na.rm = TRUE)
exclusion_condition <- exclusion_condition %>% select(cap_id, exclusion_dx, exclusion_dx_date) %>% distinct()

exclusion_patients <- bind_rows(
  exclusion_diagnosis,
  exclusion_condition
) %>%
  filter(
    !is.na(cap_id),
    !is.na(exclusion_dx),
    !is.na(exclusion_dx_date)
  ) %>%
  arrange(cap_id, exclusion_dx_date) %>%
  group_by(cap_id) %>%
  slice(1) %>%
  ungroup()

head(exclusion_patients)
```

## Final hypothyroid dx cohort, no exclusion dx

```{r}
`%not_in%` = Negate(`%in%`)
hypothyroidism_patients <- hypothyroidism_patients[hypothyroidism_patients$cap_id %not_in% exclusion_patients$cap_id,]
```

```{r}
demographics <- read_csv(
  file.path(capricorn_dir, "demographics.csv"),
  col_select = c(
    STUDY_ID,
    AGE_AT_FIRST_ENCOUNTER,
    SEX,
    HISPANIC,
    RACE
  ),
  show_col_types = FALSE
) %>%
  transmute(
    cap_id = as.character(STUDY_ID),
    age_enc = AGE_AT_FIRST_ENCOUNTER,
    sex = SEX,
    race = RACE,
    ethnicity = HISPANIC
  ) %>%
  distinct(cap_id, .keep_all = TRUE)

head(demographics)
```

## Eligible Patient Dataset

```{r}
dataset_1_eligible_patients <- hypothyroidism_patients %>%
  inner_join(
    demographics,
    by = "cap_id"
  ) %>%
  filter(
    !is.na(age_enc),
    age_enc >= 18
  ) %>%
  select(
    cap_id,
    hypo_dx,
    hypo_dx_date,
    age_enc,
    sex,
    race,
    ethnicity
  ) %>%
  arrange(cap_id)

head(dataset_1_eligible_patients)
```

## Checks and Export

```{r}
dataset_1_eligible_patients %>%
  summarise(
    rows = n(),
    unique_patients = n_distinct(cap_id),
    missing_dx_date = sum(is.na(hypo_dx_date)),
    missing_age = sum(is.na(age_enc))
  )
```

```{r}
saveRDS(dataset_1_eligible_patients, paste0(output_folder, "dataset_1_eligible_patients.Rds"))
```
